---
title: "Data Types in Python"
subtitle: "Session 2, MY470"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_float:
      collapsed: true
    toc_depth: 4
    theme: lumen
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{=html}
<style>
  body .main-container {
    max-width: 1100px;
    font-size: 12pt;
  }
</style>
```
[MY470 Homepage](https://kevinli03.github.io/notes/#MY470_Computer_Programming)

Week 2, MY470 Computer Programming

-   Title: Data Types, Operators, Methods

-   Topics: Data Types

-   Readings: N/A

<br />

------------------------------------------------------------------------

[MY470 Homepage](https://kevinli03.github.io/notes/#MY470_Computer_Programming)

# Key Points

[Scalar Data Types]{.underline} in Python

```{python}
# int - whole numbers
integer1 = 23

# float - decimals
float1 = 23.2342

# bool - True/False
bool1 = True
bool2 = False

#Empty/nothing data type
empty = None
```

<br />

[**Non-Scalar Data Types**]{.underline} in Python

```{python}

# lists - ordered and changeable
list1 = [1,2,3,4]

# tuple - ordered but unchangeable
tuple1 = (1,2,3,4)

#string - ordered group of characters
string1 = 'Hello World!'

# set - unordered, only unique objects
set1 = {1, 2, 3, 4}

# dictionary = unordered, unique objects, but with "key" for each
dict1 = {'key1' : 1, 'name2' : 2}
```

<br />

We can [**change**]{.underline} data types by using a function of that data type on something else. Be aware, data may be altered/broken.

```{python}
# let us change a string into an integer
a = int('452')
type(a)
```

<br />

Some select [**operators**]{.underline} (most common ones are obvious):

```{python}
# string concatenate (only works for strings)
'hel' + 'lo!'

# length (for any non-scalar)
len('hello!')

#in (for any non-scalar)
'f' in 'father'
```

<br />

Objects have [**methods**]{.underline} associated with them

-   A method is something you can do to a particular data type (methods are specific to the particular data type)

-   We attach a method to an object by using a . followed by the method, both after the object: object.method()

For example, for string data, you can turn all the characters upper case with the method upper()

```{python}
string1 = 'hello'
string1.upper()
```

<br />

[**Indexing**]{.underline} is possible for strings/tuples/lists, and dictionaries.

-   Strings/tuples/lists are indexed by the element order (start counting from 0).

    -   To index from the back, use a "-" sign

    -   To index a range of elements, use : between the range.

-   Dictionaries indexed by the name of the key

```{python}
# 3rd element of string/tuple/list
'London'[2]

# indexing dictionary using key
dict2 = {'Kevin': 'politician', 'Daphne': 'Biologist', 'Janice': 'Violinist'}
dict2['Daphne']
```

<br />

------------------------------------------------------------------------

[MY470 Homepage](https://kevinli03.github.io/notes/#MY470_Computer_Programming)

# Python

### Python Programming Language

Why are we learning python?

-   Python is open source: it is freely available and well documented

-   Simple and concise syntax that is intuitive to learn

-   Many useful libraries to install for practical applications

-   Widely used in industry and science - it is arguably the most widely used language today

<br />

### Python Syntax

Python has a very simple syntax. For example, for "hello world", python is simply:

```{python}
print('Hello World!')
```

<br />

Compare this to Java, which takes the following syntax:

```{r, eval = FALSE}

public class HelloWorld {
  public static void main (String[] args) {
    System.out.println("Hello world!");
  }
}
```

In comparison, Python's simple syntax makes it very intuitive to learn

<br />

### Speed of Python

Python, while simple, is not the fastest language

-   Languages that are lower level (C, Java) are far more faster

-   Python is a much higher language - so more layers between it and the hardware

However, compared to R, python is faster

-   This is not even including R packages

While there are many advantages to Python, it is not great if speed is your number one priority

<br />

### Brief History of Python

Python was started in 1989 by Guido Van Rossum

-   Python 2.0 was released in 2000, although today, it is not common

-   Nowadays, Python 3.0 is the most common version

-   Newer versions are often not backwards compatible with older verisions

<br />

------------------------------------------------------------------------

[MY470 Homepage](https://kevinli03.github.io/notes/#MY470_Computer_Programming)

# Scalar Data Types

### Introduction

There are several types of scalar (indivisible) data types

-   Integers - just a whole number, negative or positive

-   Float: similar to integer, but has decimals (distinction due to memory storing)

-   Boolean: True/False (With one capital letter)

-   NoneType: defined as "None", a placeholder to intialise a variable and later you can refer to it

Note: String is non-scalar in python (also, use single quote marks '' to denote them)

To check the type of a variable, use the function **type()**

```{python}
type(2383.34)
```

<br />

### Converting Between Types

You can convert between scalar data types

-   For example, store an integer with a float

For example, 123 is normally an integer, but using the name of the data type, we can make it a float. we can also test others

```{python}
type(float(123))

type(bool(123))

type(str(123))
```

Some notes regarding this:

-   If you change a decimal to an integer, it just drops the decimal

-   Every object has a boolean number - every number is true except 0, and for any string, it is true except for the empty string

Be careful with changing the types - this can result in errors and mistakes

<br />

This is quite useful - let us say you have a dataset

-   But, some variables, for some reason, are labelled as strings even though they are numbers

-   With this, we can quickly make them into flaots

<br />

------------------------------------------------------------------------

[MY470 Homepage](https://kevinli03.github.io/notes/#MY470_Computer_Programming)

# Non-Scalar Data Types

### List, Tuple, and Strings

[List]{.underline}: a mutable ordered sequence of values

-   Order matters - it is preserved

-   List is able to replace/add items without overriding the previous lists (useful for if you want to keep updating by adding values)

We can make a list with the square brakets [], as following:

```{python}
a = [1,3,5,7]
type(a)
```

<br />

[Tuple]{.underline}: a immutable ordered sequence of values

-   Order matters - it is preserved

-   Tuple - you cannot replace/add items, you have to override (so useful for things that aren't going to change)

-   Then why tuple? It is more memory efficient and faster, and also, they can be used for dictionary keys (see below)

We can make a tuple with round brackes (), as following:

```{python}
b = (1,2,3,4)
type(b)
```

<br />

### Strings

[String]{.underline}: a immutable ordered sequence of characters

-   Order matters in a string

-   You cannot replace/add items within the string, you have to override

We can make a string with single quotes '', as following:

```{python}
c = 'hello'
type(c)
```

<br />

Strings actually can be defined with single quotes '', or double quotes ""

-   Normally, this makes no difference

-   But, if we want to embed a certain type of quote within our string, we can use the other quote to start/end the string

For example, let us take this example:

```{python}
string = 'I am your "father"'
print(string)
```

You could also use the backslash \\ to include quotes (and other symbols)

```{python}
string = "I am your \"father\""
print(string)
```

<br />

### Sets and Dictionaries

[Set:]{.underline} A mutable unordered collection of unique values

-   Order does not matter

-   It only saves unique values

For example, we can make the string 'hello' into a set. Notice how the order of a set doesn't matter, and it gets rid of the duplicate "l".

```{python}
d = set('hello')
print(d)
```

You can define sets as well with curly brackets {}. Sets can also have different data types:

```{python}
e = {3, 'f', None}
type(e)
```

<br />

[Dictionary]{.underline} is a collection of key/value pairs

-   For example, a phone book - name and phone number

-   We have a key (ex. a name), and a value

-   It is basically a set, but with a key for each value/object

For example, let us create a dictionary with {}, with ":" assigning a key to an object:

```{python}
f = {'object1' : 1, 'coolobject' : [1,2,3]}
type(f)
```

We can also create a dictionary as follows:

-   Basically, within dict(), we have a list [] with 2 elements - both being tuples, and each tuple is a key/value combination

-   Keys of a dictionary must be a tuple

```{python}
g = dict( [ ('key1', 'value 1'), ('key2', 'value2') ] )
print(g)
```

<br />

------------------------------------------------------------------------

[MY470 Homepage](https://kevinli03.github.io/notes/#MY470_Computer_Programming)

# Operators

### Arithmetic Operators

Arithmetic operators are used to do mathematical operations to modify integers and floats:

-   $+$ addition (note - doing this on non-integers/floats will do something else)

-   $-$ subtraction

-   $*$ multiplication (note - doing this on non-integers/floats will do something else)

-   $/$ division

-   $\%$ modulus (remainder of a division problem)\\

-   $//$ floor division (division but getting rid of the decimal in the answer)

-   $**$ exponent

For example, we can do the following arithmetic:

```{python}
8339 + 5
```

<br />

However, like I mentioned before, some of the operators will function differently given a non-interger/float data type.

-   For example, if we do the same operation, but make both our number strings, it actually concatenates (adds 5 to the end of the string)

```{python}
'8339' + '5'
```

<br />

### Boolean Operators

There are 3 main boolean operators:

-   $and$ - when 2 statements both have to be met to return a True

-   $or$ - when either statement has to be met to return a True

-   $not$ - basically the opposite

For example, the most basic boolean example is:

```{python}
not False
```

These are very useful for conditional if loops

<br />

### Comparison Operators

Comparison operators allows us to compare statements, which will return a Boolean value (True/False):

-   $==$- equals (make sure it is double equal sign, since the assignment operator is one equal sign)

-   $!=$ - does not equal

-   $>$ greater than

-   $<=$ less than or equal to

For example, take this statement:

```{python}
5 == 4
```

<br />

### Assignment Operators

Assignment operators allow us to define variables

-   $=$ assigns the right side, to the left side

-   $+=$ adds right to the left side, then assigns it to the left side

-   $-=$ subtracts right from left side, then assigns to the left side

For example, we can assign a value to a variable:

```{python}
myvar = 3

#now, does 3 = 4?
myvar == 4

#take 1, add it to myvar, then store in myvar
myvar += 1

#now, does it equal 4?
myvar == 4
```

<br />

### Membership Operator

A membership operator is primarily used in non-scalar objects

-   it is defined by the syntax $in$

-   Basically, it tells us True/False that a certain element is within the non-scalar object

For example, let us ask if the string 'f' is in 'father'. The output will be True/False

```{python}
'f' in 'father'
```

<br />

### Length Operator

Length operator returns the length of a non-scalar objects

-   For a string, it will give us how many characters there are

-   For a set, it will give how many unique elements it will have

-   For a dictionary, it will give how many key/value pairs there are

Length is executed with the len() function:

```{python}
len('father')
```

<br />

### Sequence Operations

If we use $*$ for lists/tuples/strings and a number, it concatenates it a number of times. For example:

```{python}
3 * (1,2)
```

You can concatenate lists and touples with $+$

```{python}
[1, 2, 3] + [4, 5, 6]
```

<br />

------------------------------------------------------------------------

[MY470 Homepage](https://kevinli03.github.io/notes/#MY470_Computer_Programming)

# Methods

### Introduction

Objects have methods associated with them

-   A method is something you can do to a particular data type (methods are specific to the particular data type)

-   We attach a method to an object by using a . followed by the method, both after the object: object.method()

For example, for string data, you can turn all the characters upper case with the method upper()

```{python}
string1 = 'hello'
string1.upper()
```

<br />

### String Methods

There are 4 most common string methods:

1.  S.upper() - change everything to upper case
2.  S.lower() - change everything to lower case
3.  S.capitalize() - capitalises the first word
4.  S.find( $x$ ) - returns the "index" of the first instance of the string input $x$
5.  S.replace( $x_1$, $x_2$) - finds string $x_1$, replaces with $x_2$
6.  S.strip() - will remove any "whitespace character" on the left and right, which includes spaces, emojis, latex characters, etc.
7.  S.split() - splits the string into words by spaces
8.  S.join( $x$ ) - combines a list of strings $x$, with each different string being joined together by the S. string involved

For example, S.split() and S.join

```{python}
string2 = 'hi my name is kevin'
# split each word into a list
list2 = string2.split()

# join each word together with '-'
joinstring = '-'
joinstring.join(list2)
```

<br />

### Set Methods

Set methods are very useful, since sets are unordered

1.  S1.union(S2) - all elements of both S1 and S2 combined
2.  S1.intersection(S2) - elements in both S1 and S2 (overlap)
3.  S1.difference(S2) - elements in S1 but not in S2
4.  S1.symmetric_difference(S2) - elements in S1 or S2, but not in both

These reflect set theory operations quite closely

<br />

For example, what are all the letters in the strings 'hello' and 'world'?

```{python}
set1 = set('hello')
set2 = set('world')
set1.union(set2)
```

<br />

### Dictionary Indexing

Dictionaries are indexed by keys - which each value has

We can find values, based on the key

For example, let us have this dictionary - with keys being names, and values being professions:

```{python}
mydictionary = {'Kevin': 'politician', 'Daphne': 'Biologist', 'Janice': 'Violinist'}
```

We can index using square brackets []. We can find the value (profession) of anyone using their key (name):

```{python}
mydictionary['Daphne']
```

<br />

### Ordered Sequence Indexing

Ordered objects, such as lists, tuples, and strings, each element has an index

-   So the first element is index $0$, 2nd element is index $1$, 3rd element is index $3$

We can index with square brackets[]. For example, what is the 3rd letter of 'London'?

```{python}
'London'[2] #2 because we start from 0
```

We can also index from the back using a negative sign. For example, what is the 2nd to last letter in London? (note, from the back, we start counting from -1)

```{python}
'London'[-2]
```

<br />

We can also index a few different elements at the same time, using ":". Note, the last index you include is not included.

For example, what are the first 3 letters of london?

```{python}
'London'[0:3] # index 3 (4th letter) not included
```

You can also include everything from the start by not putting anything before :, and include everything to the end by not putting anything after :

We can also change the "steps" while indexing over a range. For example, if I index the first 4 letters of "london", but I skip one letter every other letter (step = 2), we get:

```{python}
'London'[0:4:2]
```

<br />

### List Methods (Mutability)

We know that lists and sets and dictionaries are mutable

-   This means we can add into these non-scalar objects

-   For example, if we have a list with 3 objects, we can add a 4th object to it while keeping the other objects.

For lists, these methods allow us to use mutability:

1.  L.append( $e$) - appends a new value $e$
2.  L.insert( $i$, $e$ ) inserts at index $i$ a new value $e$
3.  L.remove( $e$ ) removes element $e$ (if repeated, will remove the first)
4.  L.extend( $x$ ), extending a list with another list $x$
5.  L.pop( $i$ ) given the index $i$, it will remove that element and give you that element's value
6.  L.sort() - if it is numerical, goes from small to large, if string, alphabetical
7.  L.reverse() - reverse sort opposite order of above

<br />

------------------------------------------------------------------------

[MY470 Homepage](https://kevinli03.github.io/notes/#MY470_Computer_Programming)

# Exercises

### Exercise 1

How many unique letters are in the sentence: "jackie will budget for the most expensive zoology equipment"

<br />

**Solution**:

First, let us assign this string to a variable:

```{python}
a = 'jackie will budget for the most expensive zoology equipment'
```

We know that sets only retain unique elements of a string. So, let us convert the variable to a set

```{python}
set(a)
```

Notice how we have one space that isn't a letter. So, let us find the length of the set, then minus 1:

```{python}
len(set(a)) - 1
```

<br />

### Exercise 2

Reformat this sentence so it is more properly formatted

```{python}
listwords = [ '   ThiS', 'sEnteNce', 'NEEDS', 'betteR!', 'formaTting']
```

<br />

**Solution:**

First, let us join this into one string

```{python}
string3 = ' '.join(listwords)
print(string3)
```

Now, let us get rid of initial spaces before ThiS

```{python}
string3 = string3.strip()
print(string3)
```

Now, let us make only the first letter of the sentence capitalised

```{python}
string3 = string3.capitalize()
print(string3)
```

Now, let us get rid of the random ! in the middle:

```{python}
string3 = string3.replace('!', '')
print(string3)
```

Now, let us add the period:

```{python}
string3 = string3 + '.'
print(string3)
```

<br />

We could also solve this in one long command:

```{python}
' '.join(listwords).strip().capitalize().replace('!', '') + '.'
```
